A demonstration of the \LaTeX class file for Statistics in Medicine with Rmarkdown

### 2021-05-17 Manuscript/main.tex
Edit Introduction

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\begin{itemize}
\tightlist
\item
  What is the problem
\item
  Previous solutions
\item
  Our method
\item
  The implication of our method
\item
  Structure of the rest of the paper
\end{itemize}

@@

\[
\begin{aligned}
p(\boldsymbol{\beta }| \boldsymbol{\gamma}) &\propto \prod\limits_{j=1}^{p}\left[{S^0_j}^{-1/2}\exp(-1/2({\beta^0_{j}}^2/S^0_{j}))\prod\limits_{k=1}^{K_j}S_{j}^{-1/2}\exp(-1/2(\beta_{jk}^2/S_{j}))\right],%\\
% p(\bs\gamma | \bs \theta) & = \prod\limits_{j=1}^{p}\prod\limits_{k=1}^{K_j} \theta_j^{\gamma_{jk}} (1-\theta_j)^{1-\gamma_{jk}}.
\end{aligned}
\] and \[
\begin{aligned}
E(Q_1) &= \log p(\textbf{y}|\boldsymbol{\beta}, \phi) + \log p(\phi) + \sum\limits_{j=1}^p\left[E({S^0_j}^{-1}){\beta^0_j}^2+\sum\limits_{k=1}^{K_j}E(S^{-1}_{j})\beta_{jk}^2\right]%\\
%E(Q_2) &= \sum\limits_{j=1}^{p}\sum\limits_{k=1}^{K_j} \left[ E(\gamma_{jk})\log \theta_j + (1-E(\gamma_{jk})) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j)
\end{aligned}
\] where \(S_{j}^0 = (1-\gamma^{0}_{j}) s_0 + \gamma^{0}_{j} s_1\) and
\(S_{j} = (1-\gamma^{pen}_{j}) s_0 + \gamma^{pen}_{j} s_1\). To
calculate two unknown quantities \(E({S_j^0}^{-1})\) and
\(E(S^{-1}_j)\), the posterior probability
\(p^0_{j} \equiv p(\gamma^{0}_{j}=1|\Theta^{t-1})\) and
\(p_{j} \equiv p(\gamma^{pen}_{j}=1|\Theta^{t-1})\) are necessary, which
can be derived via Bayes' theorem. The calculation for \(p_j\) is
slightly different from that of \(p^0_j\), as \(p_j\) depends on the
value of \(\beta_{jk}\) for \(k=1, \dots, K_j\) and \(p^0_j\) only
depends on \(\beta_j^0\). The calculation follows the equations below.

@@

With \(E({S^0}^{-1}_{j})\) and \(E(S^{-1}_{j})\) known, \(E(Q_1)\) can
be also viewed as a \(l_2\) penalized likelihood function:
\(E({S^0}^{-1}_{j})\) controls the shrinkage of the linear effect,
informing if the predictor should be included in the model;
\(E(S^{-1}_{j})\) controls the smoothness of the spline function.
\(l_2\) penalized likelihood function can be maximize using the IWLS
algorithm. The likelihood function can be approximate by a weighted
normal likelihood: \[
  p(\textbf{y}|\boldsymbol{\beta}, \phi) \approx Normal(\textbf{z}|X\beta, \phi\Sigma)
\] where the `normal response' \(z_i\) and `weight' \(w_i\) are called
the pseudo-response and pseudo-weight, respectively.The pseudo data and
the pseudo-weight are calculated by \[
    \begin{aligned}
    z_i &= \hat\eta_i - \frac{L^{'}((y_i|\hat\eta_i))}{L^{''}((y_i|\hat\eta_i))}& w_i &= - L^{''}((y_i|\hat\eta_i)),
    \end{aligned}
    \] where \(\eta_i = \sum\limits_{j=1}^p \hat f_j^{(t)}(x_{ij})\) is
the estimated linear predictor using current estimated value of
\(\boldsymbol{\beta}\), \(L^{'}\) and \(L^{''}\) are the first
derivative and second derivative of the log-likelihood function. The
\(l_2\) penalty, \$
\sum\limits\_\{j=1\}\^{}p\left[E({S^0_j}^{-1}){\beta^0_j}^2+\sum\limits_{k=1}^{K_j}E(S^{-1}_{j})\beta_{jk}^2\right]\$
can be seems as a multivariate normal distribution of mean
\(\textbf{0}\) and variance-covariance matrix
\(\Sigma(E({S^0}_1^{-1}), E(S_1^{-1}), \dots, E({S_p^0}^{-1}), E(S_p^{-1})\),
conditioning on the value of \(E({S^0_1}^{-1}), \dots E({S^0_p}^{-1})\).
An augmentation step combines the normal approximation and multivariate
normal for \(\boldsymbol{\beta}\), similarly to Ridge regression model
fitting. The estimates of \(\boldsymbol{\beta}\) and \(\phi\) can be
quickly updated in the IRLS algorithm.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

>>>>

% Pandoc citation processing

@@

\begin{itemize}
\item
  What is the problem

  \begin{itemize}
  \tightlist
  \item
    what is the GAM
  \item
    expanding GAM in to high-dimensional frame work
  \item
    Why this is important
  \item
    Why this is hard
  \end{itemize}
\item
  Previous solutions

  \begin{itemize}
  \tightlist
  \item
    What are the methods
  \item
    What are their problems
  \end{itemize}
\item
  Introduce the wiggliness matrix
\item
  Our method
\item
  The implication of our method
\item
  Structure of the rest of the paper
\end{itemize}

Nonlinear effect modeling has long history. Early work of GAMs can date
back to 1979 when \citet{Cleveland1979} proposed local smoothing for
bivariate analysis. Many mathematical and statistical models has been
proposed since, see xxx{[}\^{} TODO: insert an review citation{]}.

Many machine learning algorithms can also model nonlinear effect.
However, a wide range of machine learning models are confined by the
``black box'' nature, rendering little help on understanding the
etiology of diseases. Semi-parametric and non-parametric regression
models, are more popular thanks to its interpretability and flexibility.
Among those, generalized additive model (GAM) proposed in the seminal
work of \citet{Hastie1987} gains the most popularity thanks to its
flexibility and easiness to incorporate many different smoothing
functions. The backfitting algorithm estimates each smooth function of
an additive model iteratively using partial residuals. The backfitting
algorithm is compatible with almost any smoothing or modeling technique
to be used as the nonparametric components. However, it is unclear how
to estimate the degree of smoothness when necessary, for example for
smoothing splines. The computation cost for cross-validating the smooth
parameters can be expensive. \citet[PP.318-320]{Wood2017} provides more
discussion on the problem.

Among all the choices of smoothing function, one particular family stood
out, splines. Splines are piece-wise polynomial functions to interpolate
nonlinear signals. Most commonly used spline is cubic spline, thanks to
its smoothness property. \citep[PP.197]{Wood2017} Early spline research
focused on regression splines, where number and location of knots are
pre-determined and ordinary least squares is used to estimate the smooth
function. However, the specification of knots is not a trivial task, as
the the positioning of knots are influential on the performance of the
model. As an improvement, researchers proposed smoothing splines where
as many knots are used and later shrunk using penalized algorithms. The
smoothing spline concept was first introduced
\citep{OSullivan1986, OSullivan1988}, and later generalized and made
famous by \citet{Eilers1996}, called P-spline. Independently,
\citet{Ruppert1999} proposed similar idea. Later, Bayesian GAMs were
introduced as an alternative to address measurement error in covariates
\citep{Berry2002} or local adaptive smoothing \citep{Lang2004}.

Generalized additive model (GAM) was first proposed by the seminal work:
a response variable, \(Y\), given a \(p\)-dimensional vector of
covariates \(\boldsymbol{x}\) can be expressed as the summation of
smooth functions of \(\boldsymbol{x}\), written as \[
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),\qquad E[f_j(x_j)] = 0.
\] where \(g^{-1}\) is the inverse function of a smooth monotonic link
function. Hence, GAM solve the model fitting problem for many
distributional outcomes, such as Poisson, binomial, gamma and normal
models. Since the appearance of GAMs, it has been applied in many area
of research, such as xxx, xxx .

In many area of research, high dimensional analysis becomes predominant,
for example sequencing data analysis, electric health record data
analysis. Previous analysis heavily relies on linear models. Recently,
there are more focus putting on modeling the non-linear relationships in
high dimensional data analysis. Hence, developing more technically
matured models is imminent. Nevertheless expanding the GAM framework for
high-dimensional applications is not a trivial task. The difficulties
mainly come in two-fold: grouping nature of the basis functions, and the
balance between spline smoothness and variable sparsity. For GAM models,
a smoothing function is consists of multiple basis functions. The basis
functions of a variable is inherently grouped. The grouping structure
should be properly taken care of for accurate estimation. Moreover,
shrinkage is previously used solely for sparsity of the model in high
dimensional data analysis, or smoothness in GAM models. How to
organically combine the two together remains as a troubling question.

\hypertarget{spike-and-slab-prior}{%
\paragraph{Spike-and-slab prior}\label{spike-and-slab-prior}}

The term ``spike-and-slab'' was first coined in \citet{Mitchell1988}
describing the shape of the prior. It is a mixture distribution,
comprised with a skinny spike density \(f_{spike}(\cdot)\) for weak
signals and a flat slab density \(f_{slab}(\cdot)\) for large signals,
mathematically \[
 \beta_i|\gamma_i \sim (1-\gamma_i)f_{spike}(\beta_i) + \gamma_if_{slab}(\beta_i).
\]

The most distinct feature of the SS priors family is that it is
conditioned on a latent binary variable \(\gamma_i\) that indicates
whether the covariate \(x_i\) is included in the model. There are
various spike-and-slab priors depending on the choice for the spike
density \(f_{spike}(\cdot)\) and the slab density \(f_{slab}(\cdot)\),
for example \citet{George1993}; \citet{George1997}; \citet{Chipman1996}
for grouped variables; \citet{Brown1998} for multivariate outcomes;
\citet{Ishwaran2005}; \citet{Clyde2004} and reference therein. Here we
introduce the most popular spike-and-slab prior introduced in
\citet{George1993}, as stochastic search variable selection (SSVS).
Given a regression problem \[
Y = \boldsymbol{X}\beta+\epsilon, 
\] we can assume the prior distribution of the regression parameters
\(\beta\) depends on a vector of binary variables
\(\gamma=(\gamma_1, \dots \gamma_p)^\prime\) as follows: given
\(\gamma_i\), \(\beta_i\) are independent distributed as a mixture
priors \[
\beta_i | \gamma_i \sim (1-\gamma_i)N(0, \sigma^2 v_0) + \gamma_iN(0,\sigma^2 v_1).
\] The prior of \(\gamma\) is traditionally assumed to be identically
independent Bernoulli distribution. Even the conjugate normal prior of
\(\boldsymbol{\beta}\) provides analytic simplicity, there is no closed
form solution for the posterior density of \(\boldsymbol{\gamma}\).
\citet{George1993} described a Gibbs sampling algorithm to approximate
the posterior distribution of \(\boldsymbol{\gamma}\) for variable
selection. The same authors later compared SSVS with other Bayesian
variable selection approaches, providing some empirical suggestions on
the advantages and disadvantages of various formulations and estimation
algorithms. \citep{George1997} Among the various problem SSVS addresses,
high-dimensionality is not one of them, due to the intensive computation
needs of MCMC algorithms. \citet{George1997} suggested MCMC algorithms
works well for medium size of predictors (p=25), which is not feasible
for high-dimensional data, where the number of predictors easily exceeds
100.

To alleviate the computation burden, many methodology development
studies focus on the maximum a posteriori (MAP) estimate
\citep{Tipping2004}. EM-based algorithms are prevalent used to speed the
computation. \citet{Rockova2014a} proposed an EM-based algorithm for a
spike-and-slab Gaussian mixture prior, EMVS. The coefficient \(\beta_j\)
independently follows a mixture normal distribution
\(N(0, \sigma^2((1-\gamma_i)v_0 + \gamma_i v_1)\) for
\(0 \leq v_0 < v_1\). The parameter \(\sigma^2\) follows a inverse gamma
prior, \(IG(v/2, v\lambda /2)\). Like in the spike-and slab family, the
latent binary variables \(\boldsymbol{\gamma}\) follows a Bernoulli
prior with a hyper prior \(\theta\) follows a beta distribution, or
simpler the uniform (0,1). In the E step, the latent variable are
treated as the missing data, and calculate the posterior mean of
\(\gamma\); in the M step, a ridge estimator was used to update the
coefficients, and \(\sigma^2, \theta\) are updated accordingly. As the
ridge regression creates small but nonzero coefficients, a further step
is needed to select the variable: EMVS further suggests to threshold the
inclusion probability for variable selection or generate regularization
plot with different values of the spike scale parameter. EMVS is also
compatible with structured prior when dealing with grouped variables.
The group structure can be imposed on the prior of the latent binary
variable \(\gamma\) via either logistic regression or markov random
field.

The same authors \citep{Rockova2018b, Rockova2018} further extended the
continuous spike-and-slab Gaussian prior to the spike-and-slab double
exponential distribution, as known as spike-and-slab lasso (SSL). The
authors also lay the theoretical discussion of comparing spike-and-slab
from the penalty perspective and the Bayesian perspective. The SSL also
mitigated the problem of EMVS that the weak signals are not shrink to
zero. Yi and his group incorporated the SSL prior in the EM-IWLS
algorithm and proposed another EM algorithm, EM-cyclic coordinate
descent, based on the Lasso penalty SSL exhibit. Such application were
used in generalized linear models\citep{Tang2017a}, cox models
\citep{Tang2017} and their grouped
counterparts\citep{Tang2018, Tang2019}. Both EM based algorithm provide
deterministic solutions, which becomes a more popular property as
reproducible research becomes more valued. However, while the EMVS are
not able to provide inference of the coefficients easily, the EM-IWLS
can easily extract the variance-covariance matrix due to the IWLS
algorithm. \citet{Bai2020} provides an up-to-date summary of SSL methods
and theoretical discussion.

To the best of our knowledge, there are few GAM models proposed for high
dimensional application. In common practice, model matrix of spline
terms are generated using base functions of preference, and then feed to
a penalized model. Without proper handle of the grouping nature of bases
functions, the estimation of the smooth function is less accurate.
\citet{Bai2020Spline} shows inferior performance of such practice
compared to using an ad hoc high-dimensional additive model.

Among the specialized high-dimensional spline methods,
\citet{Ravikumar2009} extended the original backfitting algorithm
\citep{Hastie1987} to accommodate the high-dimensional setting. The
sparsity and smoothness was handled separately with a soft threshold.
The authors also demonstrated its similarity to Group Lasso.
\citet{Meier2009} proposed an a sparsity-smoothness penalty,

\[
J(f_j) = \lambda_1 \sqrt{||f_j||^2_n + \lambda_2 \int(f^{''}_j(x))^2dx},
\] which can re-formulated as a group lasso problem as well. However,
the model can not estimate the smoothness of the spline functions
correctly, as the smoothness of all spline models are controlled by the
lambda\_2. The inflexible of the model creates problems in prediction.

From the Bayesian perspective, recent works that expands to the
high-dimensional settings includes \citet{Scheipl2012} and
\citet{Bai2020Spline}. \citet{Scheipl2012} proposed an new
spike-and-slab structure prior, normal-mixture-of-inverse gamma for
semi-parametric regression using MCMC. The normal-mixture-of-inverse
gamma prior is mixture of t-distribution with an additional coefficients
mixing vector, where the vector follows a mixture of two Gaussian
distribution. \citet{Bai2020Spline} proposed an novel high dimensional
Bayesian GAM using the spike-and-slab group lasso prior. An fast fitting
EM-based algorithm is proposed. Asymptotic theory, such as posterior
contraction rates and model selection consistency are proved. Even
though, the algorithm is flexible to cover many non-Gaussian family in a
high dimensional setting, the proposed method didn't address uncertainty
quantification, i.e.~calculating credible intervals etc. However, most
of these Bayesian GAMs still suffer a common problem, computation cost.
These Bayesian GAMs uses MCMC to estimate the posterior distribution,
which would create an intense computation load.

In this dissertation, we aim to address the challenges of non-linear
signal modeling in high-dimensional application using Bayesian
Hierarchical model. Specifically, we will propose a new Bayesian
spike-and-slab prior for the generalized additive models and Cox spline
models. The contribution of the proposed prior comes in two-fold: 1) it
supports simultaneous variable selection and function selection; 2) it
provides robust estimation and prediction for the coefficients and
outcomes. As previously described, simultaneous variable selection and
its functional form selection is not a trivial problem. Nevertheless,
the Bayesian spike-and-slab prior is a perfect solution to the problem,
as the latent binary indicator can be used to for variable selection and
the spike and slab densities are used to adaptive adjust the smoothness
of the spline function. Moreover, compared to currently available
Bayesian spline models which mostly conduct a `all-in-all-out' strategy,
i.e.~either including a non-linear form of a variable or excluding the
variable completely, we offer a more flexible solution, rending one of
three possibilities: no effect of a predictor, only linear effect of a
predictor, or non-linear effect. This simplifies the process of
determining what effect a predictor have on the outcome, without
involving hypothesis testing where the degree of freedom for non-linear
effect is complicated to estimate due to the shrinkage. Moreover, out
model inherent the the robust estimation from Bayesian models.

Given the proposed model specification, the estimation of the parameters
can be easily achieved with MCMC algorithm. Nevertheless, considering
the computational burden MCMC algorithms bear especially in
high-dimensional analysis, we propose parsimonious solutions, stemming
from previous EM-IWLS \citep{Rockova2014a}and EM-coordinate descent
algorithms\citep{Tang2018, Tang2019}. The computational friendliness
allows for fitting a model with different sets of initial values,
mimicking the posterior sampling in the MCMC algorithms. We are hopeful
that this could improve the sacrificed robustness of the estimation
process due to the MAP estimation used by the EM algorithms in
comparison to the model averaging approach.

The third aim of the dissertation work will focus on the software
implementation of the proposed models and algorithms. We will implement
the proposed models and algorithms in one of the most commonly used
statistical computing environment R. Ancillary functions will be
provided in assistance to model diagnostic, interpretation and graphing.
The R implementation of the Bayesian Hierarchical additive model is
available via a freely available R package, BHAM at
(\url{https://github.com/boyiguo1/BHAM}){[}\url{https://github.com/boyiguo1/BHAM}{]}.

The rest of the paper are arranged as the followings: in Chapter 32 we
demonstrate the Bayesian Hierarchical generalized additive model, where
we provide the setup of the proposed model and two fast-fitting
algorithms. In Chapter 3, we Application to real-world datasets
demonstrates the usefulness of the proposed methodologies in Chpater 4.
Conclusion and discussions are made in Chapter 5.

@@

\[
\begin{aligned}
p(\boldsymbol{\beta }| \boldsymbol{\gamma}) &\propto \prod\limits_{j=1}^{p}\left[{S^0_j}^{-1/2}\exp(-1/2({\beta^0_{j}}^2/S^0_{j}))\prod\limits_{k=1}^{K_j}S_{j}^{-1/2}\exp(-1/2(\beta_{jk}^2/S_{j}))\right],%\\
% p(\bs\gamma | \bs \theta) & = \prod\limits_{j=1}^{p}\prod\limits_{k=1}^{K_j} \theta_j^{\gamma_{jk}} (1-\theta_j)^{1-\gamma_{jk}}.
\end{aligned}
\] and \begin{align*}
E(Q_1) &= \log p(\textbf{y}|\boldsymbol{\beta}, \phi) + \log p(\phi) + \sum\limits_{j=1}^{p} \left[E({S^0_j}^{-1}){\beta^0_j}^2+\sum\limits_{k=1}^{K_j}E(S^{-1}_{j})\beta_{jk}^2\right]\\
E(Q_2) &= \sum\limits_{j=1}^{p}\sum\limits_{k=1}^{K_j} \left[ E(\gamma_{jk})\log \theta_j + (1-E(\gamma_{jk})) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log p(\theta_j)
\end{align*} where
\(S_{j}^0 = (1-\gamma^{0}_{j}) s_0 + \gamma^{0}_{j} s_1\) and
\(S_{j} = (1-\gamma^{pen}_{j}) s_0 + \gamma^{pen}_{j} s_1\). To
calculate two unknown quantities \(E({S_j^0}^{-1})\) and
\(E(S^{-1}_j)\), the posterior probability
\(p^0_{j} \equiv p(\gamma^{0}_{j}=1|\Theta^{t-1})\) and
\(p_{j} \equiv p(\gamma^{pen}_{j}=1|\Theta^{t-1})\) are necessary, which
can be derived via Bayes' theorem. The calculation for \(p_j\) is
slightly different from that of \(p^0_j\), as \(p_j\) depends on the
value of \(\beta_{jk}\) for \(k=1, \dots, K_j\) and \(p^0_j\) only
depends on \(\beta_j^0\). The calculation follows the equations below.

@@

With \(E({S^0}^{-1}_{j})\) and \(E(S^{-1}_{j})\) known, \(E(Q_1)\) can
be also viewed as a \(l_2\) penalized likelihood function:
\(E({S^0}^{-1}_{j})\) controls the shrinkage of the linear effect,
informing if the predictor should be included in the model;
\(E(S^{-1}_{j})\) controls the smoothness of the spline function.
\(l_2\) penalized likelihood function can be maximize using the IWLS
algorithm. The likelihood function can be approximate by a weighted
normal likelihood: \[
  p(\textbf{y}|\boldsymbol{\beta}, \phi) \approx Normal(\textbf{z}|X\beta, \phi\Sigma)
\] where the `normal response' \(z_i\) and `weight' \(w_i\) are called
the pseudo-response and pseudo-weight, respectively.The pseudo data and
the pseudo-weight are calculated by \[
    \begin{aligned}
    z_i &= \hat\eta_i - \frac{L^{'}((y_i|\hat\eta_i))}{L^{''}((y_i|\hat\eta_i))}& w_i &= - L^{''}((y_i|\hat\eta_i)),
    \end{aligned}
    \] where \(\eta_i = \sum\limits_{j=1}^p \hat f_j^{(t)}(x_{ij})\) is
the estimated linear predictor using current estimated value of
\(\boldsymbol{\beta}\), \(L^{'}\) and \(L^{''}\) are the first
derivative and second derivative of the log-likelihood function. The
\(l_2\) penalty,
\(\sum\limits_{j=1}^{p}\left[E({S^0_j}^{-1}){\beta^0_j}^2+\sum\limits_{k=1}^{K_j}E(S^{-1}_{j})\beta_{jk}^2\right]\)
can be seems as a multivariate normal distribution of mean
\(\textbf{0}\) and variance-covariance matrix
\(\Sigma(E({S^0}_1^{-1}), E(S_1^{-1}), \dots, E({S_p^0}^{-1}), E(S_p^{-1})\),
conditioning on the value of \(E({S^0_1}^{-1}), \dots E({S^0_p}^{-1})\).
An augmentation step combines the normal approximation and multivariate
normal for \(\boldsymbol{\beta}\), similarly to Ridge regression model
fitting. The estimates of \(\boldsymbol{\beta}\) and \(\phi\) can be
quickly updated in the IRLS algorithm.

@@
